### 2022-03-07 [2020-08]_ACMTOMS_(MQSI)/revision/response.tex

%% General paragraph settings (no indent, big line skip).
\parindent=0pt
\parskip=9pt

%% Fonts (for title)
\font\titlebf=cmbx20
\font\titlett=cmtt20

%% Title
{\titlebf Revision of {\titlett TOMS-2021-0028} \bigskip}

%% Body
Dear reviewers,

We are deeply grateful for the comments, questions, and references you have all provided. In accordance with your feedback we have made significant changes to the manuscript to improve clarity, expand the summary of related literature, and incorporate visual comparisons with existing spline software. We hope you find these changes have resulted in an overall better manuscript.

Each numbered item and italicized text block below is a reviewer comment (or summary of similar comments). Our responses immediately follow.

{\parindent=20pt \it

\item{(1)} The authors should include references to more related works and justify the differences between this work and the many prior works on $C^2$ interpolation. [given 23 references]

}

In order to properly satisfy this request, we've added a new subsection to the introduction that summarizes the individual tracks of research represented by the conglomerate of references provided. After reviewing all of those works we are still quite certain that this manuscript represents a novel and meaningful contribution to the literature, and have justified it's position amongst those works in subsection 1.1 of the revised manuscript.

{\parindent=20pt \it

\item{(2)} The authors should include references to and comparisons with the following software packages.

\itemitem{} Costantini, P. An algorithm for computing shape-preserving interpolating splines
of arbitrary degree, Journal of Computational and Applied Mathematics, 1988, 22(1), pp. 89-136.

\item{} While the concerns about subjectivity in comparisons are acknowledged, users will still want to know the context of this work and the trade-offs would they use this package instead of other spline softwares. For this reason a comparison with other methods including visuals should be provided.

}

We appreciate this perspective and have included such a comparison in the updated Section 5. The comparison shows methods from ...

{\parindent=20pt \it

\item{()} The monotonicity of Q(x) is not proved.

}

Per Section {\bf blank}, the theory proving the monotonicity of the resulting spline is quite dense and is formally covered by the cited references Ulrich and Watson, and {\bf blank} and {\bf blank}.

{\parindent=20pt \it

\item{()} The approximation order of the algorithm is not shown.

}

A new sentence has been added to Section {\bf blank} explaining that the approximation order of the method achieves second order convergence in almost all scenarios where the approximated function is twice continuously differentiable.

{\parindent=20pt \it

\item{()} Why use piecewise degree 5? In principle, piecewise degree 3 is enough for $C^2$. The additional degrees of freedom for making the interpolant monotone could be obtained by adding knots between the breakpoints. It would be great if you could motivate your choice in more detail.

}

In general it is not possible to form a $C^2$ approximation of arbitrary data without the incorporation of additional knots. The restriction of a fixed set of knots is somewhat arbitrary, but that is the application targeted by this line of research.

{\parindent=20pt \it

\item{()} Some statements depend on two variables being ``approximately equal'', but what does that mean? Equal up to some small threshold?

}

The definition of approximately equal for this work is when two IEEE 64-bit floating point numbers are equal up to 32 of 42 mantissa bits (precision bits, or significand in base two scientific notation).

{\parindent=20pt \it

\item{()} In Algorithm 1 (and also in the text), you use the term "curvature" to refer to the second derivative. I would avoid this, since "curvature" is a term reserved for curves, but you are dealing with functions. Similarly, the term "minimum curvature derivative" is inappropriate.

}

This has been rephrased to reduce confusion.

{\parindent=20pt \it

\item{()} In Algorithm 2, it might be better to say that f is a "degree 5" polynomial, instead of "order 6".

}

Noted and changed.

{\parindent=20pt \it

\item{()} Your binary search results in a particular set of derivatives, for which the local quintic piece is guaranteed to be monotone, but it is clearly not a unique choice. Alternatively, one could try to find, among all possible sets of derivatives and corresponding monotone quintic interpolants, the best, according to some criterion, e.g. least L2 norm. This would turn the problem into a constrained optimization problem, which is probably harder to solve, but it might, at the same time, give better results and overcome the limitation shown in Fig. 1. More motivating details would be appreciated.

}

This was a point of our own severe consideration before submitting the manuscript and these comments reflect our thoughts. In proper response to this we implemented naive $L2$ minimization algorithm that uses gradient descent over the spline representation to enforce local monotonicity constraints in a Lagrangian fashion (with an adaptive multiplier on monotonicity violations). The results we achieved were quite mixed and we've incorporated our commentary from this excursion into Section 4 of the manuscript. In truth, formally expanding on this trajectory would require an entirely new manuscript with deeper study of the problem, while it is not immediately obvious that the results would be {\it better} than the present approach. Included in this response (to reviewers only) are some of the preliminary visuals we produced in exploring this approach.

{\parindent=20pt \it

\item{()} It is quite probable that a user would want to convert the piecewise Hermite representation of the interpolant to quintic B-spline form. But the authors say in Sections 3 and 4 that they solve a banded linear system of length n (n the number of data points) to make this conversion. However, this is not necessary. At each data point, only three of the quintic B-splines are non-zero. Therefore, the three associated B-spline coefficients can be computed from the value, first, and second derivatives of the interpolant at that single point. Thus one can just solve a small 3*3 linear system to get one group of 3 B-spline coefficients. Well, it requires a bit of algebraic work to write out this 3*3 linear system, but it can be done.

}

As mentioned in the manuscript, the banded linear system representation is favored for two reasons in this package. The complexity order is unchanged by solving a single banded system, while the numerical stability is {\it theoretically} improved. In practice it is almost certainly slower than solving individual linear systems, but the increased stability and decreased redundancy in representation of the current approach is favored here. For users interested in constructing alternative representations, the optional argument {\it \bf UV} exists precisely for returning the computed first and second derivatives at knots.

{\parindent=20pt \it

\item{()} I'm confused by the O(1) and O(n) discussion in Section 4.
If there are n data points then surely the main algorithm (estimating first and second
derivatives at each data point) must require at least an O(n) algorithm, not O(1).

}

This was poorly worded on our part. The entire algorithm is O(n), however the binary search itself has a fixed number of loop iterations. The original intent of the statement was that the complexity order of the outer loop (the binary search) is constant, which ensures the overall complexity of O(n) is not changed. See the new phrasing in Section {\bf BLANK} paragraph {\bf BLANK}.

{\parindent=20pt \it

\item{()} The output are values -- why not the spline coefficients? If the output are values one would hope for a visualization package.

}

Response

{\parindent=20pt \it

\item{()} I was not familiar with the term/notion  "quadratic facet model". The term facet made me believe that the paper was on bivariate interpolation. Nothing in the abstract or title prevents this misconception.

}

\bigskip

To the best of our knowledge, rejection based on a single review has not been a journal policy and we respectfully request additional reviews. We believe this algorithm and the accompanying code has great potential for TOMS, noting that the quadratic and cubic works preceding this have remained state-of-the-art for forty years. Thank you for both your time and consideration.

Sincerely,

Thomas Lux, on behalf of all authors.

\bye

%% ----------------------------------------------------------------------
%%                        Useful TeX reference.
%% 
%% https://www.math.brown.edu/johsilve/ReferenceCards/TeXRefCard.v1.5.pdf
%% 
%% ----------------------------------------------------------------------

### 2022-03-09 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
Working on revision. Added references to papers, making summaries. Initialized response document.
Added summaries of all references. Still reworking the introduction section.

\bigskip\hrule\bigskip\medskip
\heading{1. INTRODUCTION}

@@

The currently available software for monotone piecewise polynomial
interpolation includes quadratic [He and Shi 1998, McAllister and
Roulier 1981], cubic [Fritsch and Carlson 1980], and (with limited
application) quartic [Wang and Tan 2004; Piah and Unsworth 2011; Yao
and Nelson 2018] cases. In addition, a statistical method for
bootstrapping the construction of an arbitrarily smooth monotone fit
exists [Leitenstorfer and Tutz 2006], but the method does not take
advantage of known analytic properties of quintic polynomials.
Conversely the work on quadratic shape preserving splines by Schumaker
[1983] is widely utilized in theory for its conservation of both
monotonicity and convexity, but no mathematical software is published.

@@

>>>>

\heading{1. INTRODUCTION}

@@

\heading{1.1 Related Works}

[DeVore 1977] Proves approximation error bounds for monotone spline interpolation of data from monotone functions and accompanies the previous error bounds in the context of monotone polynomial interpolation of data from a monotone function.

[Beatson 1982] develops a quadratic spline curve fitting algorithm that maintains local convexity while also proving some error bounds. This method avoids inserting additional knots, but does not interpolate the data.

[Gregory and Delbourgo 1982] Proves necessary and sufficient conditions for a closed form solution to monotone C1 quadratic rational spline interpolation. Does not provide formal algorithm, nor code. Then again iterates towards a solution of nonlinear equations that ensure C2 continuity of quadratic rational splines. The solution they iterate towards exists and is unique. They provide proofs and methods for setting derivatives at the ends, no formal algorithms or code in [Delbourgo and Gregory 1983].

[Costantini 1986] Proves existence of splines that maintain monotonicity and convexity while having fixed continuity and order.

[Dougherty, Edelman, and Hyman 1989] Proposes an algorithm for constructing C2 Hermite monotone spline interpolants using proven sufficient, but not necessary conditions. As these bounds are not sharp, the space of allowable derivative and second derivative values at knots is unnecessarily shrunk. An algorithm is provided, but no code. Recommends against the direct optimization for smoothness parameters like L2, instead promoting problem specific definitions of 'geometric niceness'.

Fiorot and Tabka 1991] Proves a simple method for determining the existence of a monotone C2 cubic interpolating spline, but functions of this kind do not always exist for arbitrary convex monotone data sets.

[Delbourgo 1993] Continuous previous work by incorporating a cubic numerator, proves shape preservation when the correct value for a tension parameter is set but no method for setting that parameter. Accuracy depends on accurate derivative evaluations. No code nor formal algorithm is provided.

[Huynh 1993] Presents several nonlinear boundary conditions based on the median function to establish a fourth order monotonicity preserving constraints relying on divided difference initial derivative estimates. Overall generates a monotone C1 cubic interpolant to data.

[Pruess 1993] Proposes a cubic C2 shape preserving spline method with free parameters (mentions quintic splines are necessary for globally C2 monotone splines). Sufficient (but not necessary) conditions are provided for monotonicity.

[Manni 1997] Constructs a C2 cubic spline with two additional knots per interval. Then [Cravero and Manni 2003] Constructs C3 interpolating splines by progressively increasing tension through Bezier control polygons until shape parameters are satisfied, no code is provided.

[Abbas, Majid, and Ali 2012] No extra knots, rational functions, sufficiency and necessity proven, works for unequally spaced data, no code associated with the paper, no algorithm provided for the initialization for free shape parameters.

[Wang and Tan 2004] Propose a sufficient quartic C2 method with two free parameters.

[Piah and Unsworth 2011] improve upon the sufficient conditions of Wang and Tan for C2 quartic rational interpolation.

[Kvasov 2014] use sufficient conditions for monotonicity and convexity to form a globally weighted combination of C1 quadratic splines that satisfy shape constraints.

[Zhu and Han 2015] uses sufficient conditions over three free parameters to construct C2 rational quartic shape preserving splines. Finally [Han 2018] Proposes a method for making arbitrarily smooth shape preserving spline interpolants, but uses a centered quadratic to estimate derivatives and a sufficient set of bounding conditions to ensure shape preservation in practice. No necessary conditions are applied, no algorithm is provided, neither is any code provided.

[Yao and Nelson 2018] Develop novel sufficient conditions for monotone C2 quartic spline interpolation and maintain shape with added control points.

\heading{1.2 Available Software}

The currently available peer reviewed and published software for
monotone piecewise polynomial interpolation is severely lacking in
comparison with the number of published methodologies mentioned
above. This is indicative of how difficult it is to properly handle
the numerical conditions that arise when constructing and evaluating
precise shape preserving splines.

The piecewise quadratic $C^1$ method of [McAllister and Roulier 1981]
is available through TOMS, as well as the sufficient $C^2$ quintic
method of BVSPIS. The sufficient $C^1$ cubic spline method of [Fritsch and Carlson 1980],
is available through the SciPy Python package, and the $C^1$ rational
quadratic method of Schumaker [1983] is available as an R package.

cases. In addition, a statistical method for bootstrapping the
construction of an arbitrarily smooth monotone fit exists
[Leitenstorfer and Tutz 2006], but the method does not take advantage
of known analytic properties of quintic polynomials.  Conversely the
work on quadratic shape preserving splines by Schumaker [1983] is
widely utilized in theory for its conservation of both monotonicity
and convexity, but no mathematical software is published.

@@

\heading{1.3 Overview}

### 2022-03-09 [2020-08]_ACMTOMS_(MQSI)/paper/mqsi20.tex
Added summaries of all references. Still reworking the introduction section.

@@

%% 
\refj{Beatson, R. K.}{1982}{Monotone and convex approximation by splines: error estimates and a curve fitting algorithm}{SIAM Journal on Numerical Analysis}{19(6):1278--1285}

\refj{Cravero, I., C. Manni}{2003}{Shape-preserving interpolants with high smoothness}{Journal of computational and applied mathematics}{157, 383--405}

\refj{Delbourgo, R., Gregory,  J. A.}{1983}{$C^2$ rational quadratic spline interpolation to monotonic data}{IMA J. Numer. Anal.}{3, 141--152}

\refj{Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in tension}{SIAM J. Numer. Anal.}{30, 595--607}

\refj{DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM J. Math. Anal.}{8, 891-905}

\refj{DeVore, R. A.}{1977}{Monotone approximation by polynomials}{SIAM J. Math. Anal.}{8, 906-921}

\refj{Dougherty, R. L., Edelman, A., Hyman, J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving cubic and quintic Hermite interpolation}{Math. Comp.}{52, 471--494}

\refj{Gregory, J. A.,Delbourgo,  R.}{1982}{Piecewise rational quadratic interpolation to monotonic data}{IMA J. Numer. Anal.}{2, 123--130}

\refj{Fiorot, J. C., Tabka, J.}{1991}{Shape-preserving $C^2$ cubic polynomial interpolating splines}{Math. Comp.}{57, 291--298}

\refj{Han, X.}{2018}{Shape-preserving piecewise rational interpolation with higher order continuity}{Applied Mathematics and Computation}{337, 1-13}

\refj{Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM J. Numer. Anal.}{30, 57--100}

\refj{Kvasov, B. I.}{2014}{Monotone and convex interpolation by weighted quadratic splines}{Adv. Comput. Math.}{40, 91-116}

\refj{Manni, C., Sablonni\`ere, P.}{1997}{Monotone interpolation of order 3 by $C^2$ cubic splines}{IMA J. Numer.Anal.}{17, 305--320}

\refj{Manni, C.}{2001}{On shape preserving $C^2$ Hermite interpolation}{BIT Numerical Mathematics}{41, 127--148}

\refj{Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline interpolation}{IMA J. Numer. Anal.}{13, 493--507}

\refj{Zhu, Y.,Han, X.}{2015}{$C^2$ rational quartic interpolation spline with local shape preserving property}{Appl. Math. Lett.}{46, 57-63}

\refj{Zhu, Y.,Han, X.}{2015}{Shape preserving $C^2$ rational quartic interpolation spline with two parameters}{International Journal of Computer Mathematics}{92, 2160-2177}

@@

% haralick1981facet
\refj{Haralick, R. M., and Watson, L. T.}{1981}{A facet model for
  image data.}{Computer Graphics and Image Processing}{15(2):113--129}

% he1998monotone
\refj{He, X. and Shi, P.}{1998}{Monotone B-spline smoothing}{Journal
  of the American Statistical Association}{93(442):643--650}

>>>>

%% Proves approximation error bounds for monotone spline interpolation of data from monotone functions.
\refj{DeVore, R. A.}{1977}{Monotone approximation by splines}{SIAM Journal on Mathematical Analysis}{8(5):891--905}

%% Accompanies the previous error bounds in the context of monotone polynomial interpolation of data from a monotone function.
\refj{DeVore, R. A.}{1977}{Monotone approximation by polynomials}{SIAM Journal on Mathematical Analysis}{8(5):906--921}

%% Develops a quadratic spline curve fitting algorithm that maintains local convexity while also proving some error bounds. This method avoids inserting additional knots, but does not interpolate the data.
\refj{Beatson, R. K.}{1982}{Monotone and convex approximation by splines: error estimates and a curve fitting algorithm}{SIAM Journal on Numerical Analysis}{19(6):1278--1285}

%% Proves necessary and sufficient conditions for a closed form solution to monotone C1 quadratic rational spline interpolation. Does not provide formal algorithm, nor code.
\refj{Gregory , J. A.,Delbourgo,  R.}{1982}{Piecewise rational quadratic interpolation to monotonic data}{IMA Journal of Numerical Analysis}{2(2):123--130}

%% Iterates towards a solution of nonlinear equations that ensure C2 continuity of quadratic rational splines. The solution they iterate towards exists and is unique. They provide proofs and methods for setting derivatives at the ends, no formal algorithms or code.
\refj{Delbourgo, R., Gregory, J. A.}{1983}{$C^2$ rational quadratic spline interpolation to monotonic data}{IMA Journal of Numerical Analysis}{3(2):141--152}

%% Proposes an algorithm for constructing C2 Hermite monotone spline interpolants using proven sufficient, but not necessary conditions. As these bounds are not sharp, the space of allowable derivative and second derivative values at knots is unnecessarily shrunk. An algorithm is provided, but no code. Recommends against the direct optimization for smoothness parameters like L2, instead promoting problem specific definitions of 'geometric niceness'.
\refj{Dougherty, R. L., Edelman, A., Hyman, J. M.}{1989}{Nonnegativity-, monotonicity- convexity-preserving cubic and quintic Hermite interpolation}{Mathematics of Computation}{52(186):471--494}

%% Proves a simple method for determining the existence of a monotone C2 cubic interpolating spline, but functions of this kind do not always exist for arbitrary convex monotone data sets.
\refj{Fiorot, J. C., Tabka, J.}{1991}{Shape-preserving $C^2$ cubic polynomial interpolating splines}{Mathematics of Computation}{57(195):291--298}

%% Continuous previous work by incorporating a cubic numerator, proves shape preservation when the correct value for a tension parameter is set but no method for setting that parameter. Accuracy depends on accurate derivative evaluations. No code nor formal algorithm is provided.
\refj{Delbourgo, R.}{1993}{Accurate $C^2$ rational interpolations in tension}{SIAM Journal on Numerical Analysis}{30(2):595--607}

%% Presents several nonlinear boundary conditions based on the median function to establish a fourth order monotonicity preserving constraints relying on divided difference initial derivative estimates. Overall generates a monotone C1 cubic interpolant to data.
\refj{Huynh, H. T.}{1993}{Accurate monotone cubic interpolation}{SIAM Journal of Numerical Analysis}{30(1):57--100}

%% Proposes a cubic C2 shape preserving spline method with free parameters (mentions quintic splines are necessary for globally C2 monotone splines). Sufficient (but not necessary) conditions are provided for monotonicity.
\refj{Pruess, S.}{1993}{Shape preserving $C^2$ cubic spline interpolation}{IMA Journal of Numerical Analysis}{13(4):493--507}

%% Constructs a C2 cubic spline with two additional knots per interval. No code (paywall).
\refj{Manni, C., Sablonni\`ere, P.}{1997}{Monotone interpolation of order 3 by $C^2$ cubic splines}{IMA Journal of Numerical Analysis}{17(2):305--320}

%% Uses sufficient conditions to generate C2 cubic spline. No code (paywall).
\refj{Manni, C.}{2001}{On shape preserving $C^2$ Hermite interpolation}{BIT Numerical Mathematics}{41(1):127--148}

%% Constructs C3 interpolating splines by progressively increasing tension through Bezier control polygons until shape parameters are satisfied, no code is provided.
\refj{Cravero, I., C. Manni}{2003}{Shape-preserving interpolants with high smoothness}{Journal of computational and applied mathematics}{157(2):383--405}

@@

%% Use sufficient conditions for monotonicity and convexity to form a globally weighted combination of C1 quadratic splines that satisfy shape constraints.
\refj{Kvasov, B. I.}{2014}{Monotone and convex interpolation by weighted quadratic splines}{Advances in Computational Mathematics}{40(1):91--116}

%% Uses sufficient conditions over three free parameters to construct C2 rational quartic shape preserving splines.
\refj{Zhu, Y.,Han, X.}{2015}{$C^2$ rational quartic interpolation spline with local shape preserving property}{Applied Mathematics Letters}{46:57--63}

%% Uses sufficient conditions over two free parameters to construct C2 rational quartic shape preserving splines.
\refj{Zhu, Y.,Han, X.}{2015}{Shape preserving $C^2$ rational quartic interpolation spline with two parameters}{International Journal of Computer Mathematics}{92(10):2160--2177}

%% Proposes a method for making arbitrarily smooth shape preserving spline interpolants, but uses a centered quadratic to estimate derivatives and a sufficient set of bounding conditions to ensure shape preservation in practice. No necessary conditions are applied, no algorithm is provided, neither is any code provided.
\refj{Han, X.}{2018}{Shape-preserving piecewise rational interpolation with higher order continuity}{Applied Mathematics and Computation}{337:1--13}

@@
